---
title: "HCZ-model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HCZ-model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MetaRVM)
```

We will create some sample input for this model run. (This data sets are also provided as part of the package).


```{r}
## Model initialization
sim_length <- 100

N_pop <- 6
pop_init <- data.frame(N = rep(NA, N_pop),
                       S0 = rep(NA, N_pop),
                       I0 = rep(NA, N_pop),
                       V0 = rep(NA, N_pop))

pop_init$N <- sample(1000:1500, N_pop)
pop_init$I0 <- sample(10:100, N_pop)
pop_init$V0 <- sample(1:10, N_pop)
pop_init$R0 <- sample(100:200, N_pop)
pop_init$S0 <- pop_init$N - pop_init$I0 - pop_init$V0

pop_zones <- pop_init

## Vaccine profile
tt <- seq(0, sim_length, by = 1)
nt <- length(tt)
vol <- sample(1:10, (nt-1)*N_pop, replace = TRUE)

vac_zones <- cbind(tt, rbind(pop_zones$V0, matrix(vol, nrow = nt-1)))
colnames(vac_zones) <- c("t", paste0("v", 1:N_pop))

## mixing matrix

m1 <- m2 <- m3 <- m4 <- matrix(NA, nrow = N_pop, ncol = N_pop)
diag(m1) <- round(runif(N_pop, min = 0.85, max = 0.92), 2)
diag(m2) <- 1
diag(m3) <- round(runif(N_pop, min = 0.7, max = 0.9), 2)
diag(m4) <- 1

for(ii in 1:N_pop){
  tmp <- runif(N_pop - 1)
  m1[ii, -ii] <- round(tmp * (1 - m1[ii, ii]) / sum(tmp), 2)
  m2[ii, -ii] <- 0
  m3[ii, -ii] <- round(tmp * (1 - m3[ii, ii]) / sum(tmp), 2)
  m4[ii, -ii] <- 0
}

m_weekday_day <- m1
m_weekday_night <- m2
m_weekend_day <- m3
m_weekend_night <- m4
```


```{r}
out <- meta_sim(N_pop = N_pop,
                beta_i = 0.7,
                beta_v = 0.2,
                S0 = pop_zones$S0,
                I0 = pop_zones$I0,
                P0 = pop_zones$N,
                V0 = pop_zones$V0,
                R0 = pop_zones$R0,
                m_weekday_day = m_weekday_day,
                m_weekend_day = m_weekend_day,
                m_weekday_night = m_weekday_night,
                m_weekend_night = m_weekend_night,
                delta_t = 0.5,
                tvac = vac_zones[, 1],
                vac_mat = as.matrix(vac_zones[, 2:ncol(vac_zones)]),
                VtoS = 0.02,
                 EtoIpresymp = 0.33,
                 etopa = 0.5,
                 pretoIsymp = 0.5,
                 IasymptoR = 0.2,
                 IsymptoRH = 0.5,
                 istohr = 0.5,
                 HtoRD = 0.33,
                 htor = 0.5,
                 RtoS = 0.02,
                 vac_eff = 0.5,
                 nsteps = 100,
                 is.stoch = 0,
                 seed = NULL)

# head(out)
```

There are some utility functions to extract a subset of disease states from the output data frame. For example, if we want to extract all infectious compartments, we can do this by the utility function `get_disease_state`.

```{r}
out_I <- get_disease_state(data.frame(out), disease_states = c("I_presymp", "I_asymp", "I_symp"))
```

## Vizualizations

These are some easy way to create visualizations on the output of the simulation. These functionalities will be added later as functions to the package.

#### Aggregate level plot

```{r}
library(ggplot2)
library(dplyr)

compartment_colors <- c("S" = "steelblue3",
                              "E" = "tan1",
                              "I_presymp" = "salmon3",
                              "I_asymp" = "orangered2",
                              "I_symp" = "red4",
                              "H" = "mediumpurple",
                              "R" = "green",
                              "D" = "grey",
                              "V" = "darkgreen")

long_out <- get_disease_state(data.frame(out))

ggplot(long_out %>% filter(disease_state %in% c("S", "E", "H", "D",
                                                "I_presymp", "I_asymp",
                                                "I_symp", "R", "V")) %>%
                          mutate(disease_state = factor(disease_state,
                                                        levels = c("S", "E", "H", "D",
                                                                   "I_presymp", "I_asymp",
                                                                   "I_symp", "R", "V"))) %>%
                          group_by(step, disease_state) %>%
                          summarize(total_value = sum(value), .groups = "drop"),
       aes(x = step, y = total_value, color = disease_state)) +
          geom_line(linewidth = 0.5, alpha = 0.5) +
          scale_color_manual(values = compartment_colors) +
          labs(
            title = "Disease Compartments Over Time",
            x = "steps",
            y = "Counts",
            color = "Compartment",
          ) +
          theme_bw() +
          theme(
            plot.title = element_text(hjust = 0.5),
            legend.position = "right"
          )
```
